--- 
title: "Statistical Genetics: Analyses with R"
author: "Dr. Shirin Glander"
date: '`r Sys.Date()`'
output: pdf_document
bibliography:
- book.bib
- packages.bib
description: Description to come...
documentclass: krantz
link-citations: yes
lof: yes
lot: yes
colorlinks: yes
site: bookdown::bookdown_site
biblio-style: apalike
fontsize: 12pt
monofont: "Source Code Pro"
monofontoptions: "Scale=0.7"
---

```{r setup, include=FALSE}
options(
  htmltools.dir.version = FALSE, formatR.indent = 2, width = 55, digits = 4
)

# install the packages needed by this book; you fill out c(), e.g. c('ggplot2', 'dplyr')
lapply(c(), function(pkg) {
  if (system.file(package = pkg) == '') install.packages(pkg)
})
```

# Preface {-}

Hi there, this is my great book.

## Why read this book {-}

It is very important...

## Structure of the book {-}

Chapters \@ref(introduction) introduces a new topic, and ...

## Software information and conventions {-}

I used the **knitr**\index{knitr} package [@xie2015] and the **bookdown**\index{bookdown} package [@R-bookdown] to compile my book. My R session information is shown below:

```{r}
sessionInfo()
```

Package names are in bold text (e.g., **rmarkdown**), and inline code and filenames are formatted in a typewriter font (e.g., `knitr::knit('foo.Rmd')`). Function names are followed by parentheses (e.g., `bookdown::render_book()`).

## Acknowledgments {-}

A lot of people helped me when I was writing the book.

```{block2, type='flushright', html.tag='p'}
Frida Gomam  
on the Mars
```

