--- 
title: 'Statistical Genetics: Analyses with R'
author: "Dr. Shirin Glander"
date: '`r Sys.Date()`'
knit: "bookdown::render_book"
documentclass: krantz
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
fontsize: 12pt
monofont: "Source Code Pro"
monofontoptions: "Scale=0.7"
site: bookdown::bookdown_site
---

```{r setup, include=FALSE}
options(
  htmltools.dir.version = FALSE, formatR.indent = 2, width = 55, digits = 4
)

# install the packages needed by this book; you fill out c(), e.g. c('ggplot2', 'dplyr')
lapply(c(), function(pkg) {
  if (system.file(package = pkg) == '') install.packages(pkg)
})
```

# Preface {-}

Hi there, this is my great book.

## Why read this book {-}

It is very important...

## Structure of the book {-}

Chapters \@ref(introduction) introduces a new topic, and ...

## Software information and conventions {-}

I used the **knitr**\index{knitr} package and the **bookdown**\index{bookdown} package to compile my book. My R session information is shown below:

```{r}
sessionInfo()
```

Package names are in bold text (e.g., **rmarkdown**), and inline code and filenames are formatted in a typewriter font (e.g., `knitr::knit('foo.Rmd')`). Function names are followed by parentheses (e.g., `bookdown::render_book()`).

Some packages are available via CRAN, while others are hosted at Bioconductor. I will
provide package installation instructions at the beginning of each example, indicating where
each package can be found.
I will also be using the library() function, rather than require() for loading packages to
make sure that we will get an error message in case packages have not been installed correctly.
The example workflows included are meant to illustrate the theoretical concepts and get you
started on your own analysis. They are minimal examples of the necessary steps but are not
meant to substitute the package manuals. When you want to apply the workflows to your
own data, I highly recommend going back to the package documentation to find out about
additional functions and using the help() function to explore parameter options. I will be
using the same naming and code schemes as in the package manuals to make finding the
relevant parts easy.

## Acknowledgments {-}

A lot of people helped me when I was writing the book.

```{block2, type='flushright', html.tag='p'}
Shirin Glander
```

